### 2021-04-07 sections/2-info-theory.tex
typos, reorganization, and more

\section{Information Theoretic Properties}
\label{sec:information-theoretic-properties}

We extend the work from~\cite{berger2020levenshtein} for search and compression using a flat clustering to our case of a hierarchical clustering.
We further generalize this for any dataset and distance function described below.

@@

\begin{enumerate}[i.]
    \item $f(x, y) \mapsto \mathbb{R}$ is deterministic.
    \item $f(x, y) = 0 \Leftrightarrow x = y$ and $f(x, y) > 0 \Leftrightarrow x \neq y$, i.e. distance values are non-negative.
    \item $f(x, y) = f(y, x)$, i.e. symmetry.
    \item $f(x - z, y - z) = f(x, y)$, i.e. invariance under translation or encoding.
    \item $f(x, z) \leq f(x, y) + f(y, z)$, i.e. the triangle inequality holds.
\end{enumerate}

@@

We define a \textbf{Metric Ball} as $B(q, r) = \{ x \in X \ | \ f(q, x) \leq r \}$, i.e. a ball containing all instances in the dataset within distance $r$ from an instance $q$.
We also define a \textbf{Cluster} as $C(c, r)$ analogous to a Metric Ball with the caveat that if some clusters have overlapping volumes then any instances in that overlapping volume are assigned to only one of the clusters.
Thus $C_i \cap C_j = \phi \Leftrightarrow i \neq j$ and $\bigcup C = X$.

The \textbf{Metric Entropy} (Kolmogorov Entropy) of the dataset, denoted by $N_{\epsilon}^{ent}(X)$, is the maximum $n$ such that $\{x_1, \dots, x_n\} \subset X$ and $f(x_i, x_j) \geq \epsilon \ \forall \ i \neq j$, i.e. the maximum number of $\epsilon$-separated instances in the dataset.

@@

Clearly, $N_{2\epsilon}^{ent}(X) \leq N_{\epsilon}^{int}(X) \leq N_{\epsilon}^{ent}(X)$.
Therefore, the Metric Entropy and the Internal Covering Number of a dataset are equivalent measures.

The \textbf{Fractal Dimension} (Minkowski Dimension) of a dataset is
\begin{align*}
    d(X) &= \lim_{\epsilon \rightarrow 0} \frac{\log(N_{\epsilon}^{int}(X))}{\log(\frac{1}{\epsilon})}
\end{align*}

For a cluster, $C_i(c_i, r_i)$, its local fractal dimension is equivalently defined as
\begin{align}
    \label{eq:local-fractal-dimension}
    d(C_i) &= \log_2 \ \frac{|B(c_i, r_i)|}{|B(c_i, \frac{r_i}{2})|}
\end{align}

Given an \textbf{Internal Covering} of $X$, we can specify any $\bar{x} \in X$to within error $\epsilon$ by specifying $x_i$ such that $\bar{x} \in B(x_i, \epsilon)$.
This it takes $\mathcal{O} \big( \log N_{\epsilon}^{ent}(X) \big)$ bits of information to specify $\bar{x}$ to within error $\epsilon$.

Given two instances $x_1, x_2 \in X$, we can construct $x_2 - x_1$, a \textbf{minimal encoding} of $x_2$ in terms of $x_1$, such that the number of bits needed to specify the encoding scales linearly with $f(x_1, x_2)$.
We would need to construct a method for construction such \textit{minimal encodings} for each pairing of $(X, f)$.
See Sections~\ref{subsec:methods:compression} and~\ref{sec:datasets-and-distance-functions}.

@@

For example, given a dataset of genomic strings with the Levenshtein edit distance as the metric, let $x_1$ and $x_2$ be two strings in the dataset.
Let $x_2 - x_1$ be the list of edits to convert $x_1$ into $x_2$ as used for the Levenshtein distance between $x_1$ and $x_2$.
The length of this edit list is equal to the Levenshtein edit distance i.e. $|x_2 - x_1| = f(x_1, x_2)$.
Each such edit can be stored using $\mathcal{O}(\log m)$ bits where $m$ is the maximum length of any string in the dataset.
Thus, the total number of bits needed for the encoding is $\mathcal{O} \big( f(x_1, x_2) \big)$.

Hence, for $|X| < \inf$,
\begin{align*}
    H \big( C(c, r) \big) &\leq H(c) + \sum_{x \in C} H(x - c) \\
    &\leq H(c) + |C(c, r)| \cdot H(y - c) \\ \\
    \text{where} \ y = \argmax_{x \in C}& \ f(x, c) \text{, or equivalently} \ f(y, c) = r.
\end{align*}

@@

We have two choices for encoding the instances in a cluster:
either each instance is directly encoded in terms of the cluster center, 
or each child-cluster is encoded in terms of the child-center and the child-centers are encoded in terms of the cluster-center.

Thus the Shannon Entropy of a cluster encoding is given by
\begin{align}
    \label{eq:hierarchical-shannon-entropy}
    H(C) &\leq H(c - c_{parent}) + \min \begin{cases}
        |C| \cdot H(y - c) \\
        H(C_{left}) + H(C_{right})
    \end{cases}
\end{align}

@@

\subsection{Guaranteed Decreases in Radii}
\label{subsec:methods:radii-decrease}

We show that the radii of clusters are guaranteed to decrease after, at most, every $d$ recursive partitions.

The Manifold Hypothesis assumption indicates that the instances in the dataset follow a $d$-dimensional distribution where $d \ll D$, the dimensionality of the embedding space.
We measure the dimensionality of this low-dimensional distribution using the local fractal dimension of the dataset.
Thus, we can describe this distribution by choosing some set of $d$ mutually orthogonal axes.

Let $2r$ be the maximum pairwise distance among the instances in the dataset.
In the worst case, e.g. with the distribution filling a $d$-sphere, our chosen $d$ axes will be such that $2r$ is the maximum pairwise distance \textit{along} each axis.

The Partition algorithm will select a maximally distant pair of points to use as poles, i.e. it will choose one of the $d$ axes along which to partition the cluster into two children.
After one application of this Partition, the maximum pairwise distance along that axis will be bounded above by $r$.
The next recursive application of Partition will select another of the $d$ axes.
Thus, after at most $d$ Partitions, the maximum pairwise distance along each of the $d$ axes will be bounded above by $r$.
The overall maximum pairwise distance, \textit{i.e. not restricted to be along any one axis}, will be bounded above by $r \sqrt{2}$, for example, between two instances that are at the extrema of two different axes.

@@

@@

In the worst case, we would have a perfectly balanced clustering.
Starting with cluster $C$ and applying $d$ recursive partitions would give us $2^d$ descendants each with $|C| \cdot 2^{-d}$ instances.
If cluster $C$ has radius $r$ then each of the $2^d$ descendants will have radius $\frac{r}{\sqrt{2}}$.
In the general case, we can assume that the descendants have $\mathcal{O}(|C| \cdot 2^{-d})$ instances each.

Note that, in practice, we never see a perfectly balanced clustering.
Instead, Partition produces unbalanced trees due the varying density of sampling of the instances in different regions of the manifold.
Further, the cluster radii decrease by a factor much larger than $\frac{1}{\sqrt{2}}$ and they do so every one or two partitions rather than after $d$ partitions.

Let $T(C, d)$ be the subtree rooted at $C$ with depth $d$ and $J$ leaves.
Since we cannot partition a cluster containing only one instance, we know that $J \leq 2^d \leq |C|$.
Let the leaves be indexed with integers $j \in [1, J]$.
Let the center of each child cluster be encoded in terms of the center of its parent cluster.
Therefore, $H \big( T(C, d) \big)$, the Shannon Entropy of $T(C, d)$, is the sum of all encodings of child-centers in terms of their parent-centers..

Thus, the second case in Equation~\ref{eq:hierarchical-shannon-entropy} becomes:
\begin{align*}
    & H \big( T(C, d) \big) + \sum H(C_j) \\
    = \ & H \big( T(C, d) \big) + \sum |C_j| \cdot H(y_j - c_j)
\end{align*}

By the scaling behavior of cluster radii, as discussed in Section~\ref{subsec:methods:radii-decrease}, we know $H(y_j - c_j) \leq \frac{1}{\sqrt{2}} H(y - c)$.
Thus, we continue:
\begin{align*}
    \leq \ & H \big( T(C, d) \big) + \frac{1}{\sqrt{2}} \cdot \sum |C_j| \cdot H(y - c) \\
    = \ & H \big( T(C, d) \big) + \frac{1}{\sqrt{2}} \cdot |C| \cdot H(y - c)
\end{align*}

So we rewrite Equation~\ref{eq:hierarchical-shannon-entropy} as
\begin{align}
    \label{eq:hierarchical-shannon-entropy-2}
    H(C) &\leq H(c - c_{parent}) + \min \begin{cases}
        |C| \cdot H(y - c) \\
        |C| \cdot H(y - c) \cdot \frac{1}{\sqrt{2}} + H \big( T(C, d) \big) 
    \end{cases}
\end{align}

@@

In practice, for low depths in the tree, using subtrees to encode the instances produces great gains in the number of bits used.
As we continue deeper into the full tree, we start seeing diminishing returns for using subtrees.
Eventually, either the two cases equalize or the first cases becomes the better choice.
At such a depth in the tree, we encode the instances in a cluster using the cluster center rather than the subtree.
Notably, such a depth need not be uniform across the tree and can adapt to the sampling density in different regions of the manifold.

>>>>

\section{Information Theoretic Analysis}
\label{sec:information-theoretic-analysis}

We extend the work from~\cite{berger2020levenshtein} for search and compression using a flat clustering to our case of a hierarchical clustering.
We further generalize this for any dataset and distance function described below.
We also provide relevant algorithms for some parings of datasets and distance functions in Section~\ref{subsec:methods:compression}.

@@

\begin{enumerate}[i.]
    \item $f: X \times X \mapsto \mathbb{R}$ is deterministic.
    \item $f(x, y) = 0 \Leftrightarrow x = y$ and $f(x, y) > 0 \Leftrightarrow x \neq y$.
    \item $f(x, y) = f(y, x)$, i.e. symmetry.
    \item $f(x, z) \leq f(x, y) + f(y, z)$, i.e. the triangle inequality.
\end{enumerate}

@@

We define a \textbf{Metric Ball} as $B(q, r) = \{ x \in X \ | \ f(q, x) \leq r \}$, i.e. a ball containing all instances in the dataset within distance $r$ from an instance $q$.
We also define a \textbf{Cluster} as $C(c, r)$ analogous to a Metric Ball with the caveat that if some clusters have overlapping volumes then any instances in that overlapping volume are assigned to only one of the clusters.
Thus $C_i \cap C_j = \phi \Leftrightarrow i \neq j$ and $\bigcup C_i = X$.

The \textbf{Fractal Dimension} (Minkowski Dimension) of a dataset is
\begin{align*}
    d(X) &= \lim_{\epsilon \rightarrow 0} \frac{\log \big( N_{\epsilon}^{int}(X) \big) }{\log \big( \frac{1}{\epsilon} \big)}
\end{align*}

For a cluster, $C(c, r)$, its local fractal dimension is equivalently defined as
\begin{align}
    \label{eq:local-fractal-dimension}
    d(C) &= \log_2 \ \frac{|C(c, r)|}{|C(c, \frac{r}{2})|}
\end{align}

The \textbf{Metric Entropy} (Kolmogorov Entropy) of the dataset, denoted by $N_{\epsilon}^{ent}(X)$, is the maximum $n$ such that $\{x_1, \dots, x_n\} \subseteq X$ and $f(x_i, x_j) \geq \epsilon \ \forall \ i \neq j$, i.e. the maximum number of $\epsilon$-separated instances in the dataset.

@@

As in~\cite{berger2020levenshtein}, $N_{2\epsilon}^{ent}(X) \leq N_{\epsilon}^{int}(X) \leq N_{\epsilon}^{ent}(X)$, i.e. the Metric Entropy bounds the Internal Covering Number.
Thus, they are equivalent measures.

Given an \textbf{Internal Covering} of $X$, we can specify any $\bar{x} \in X$ to within error $\epsilon$ by specifying $x_i$ such that $\bar{x} \in B(x_i, \epsilon)$.
This it takes $\mathcal{O} \big( \log N_{\epsilon}^{ent}(X) \big)$ bits of information to specify $\bar{x}$ to within error $\epsilon$.

Given two instances $x_1, x_2 \in X$, we can construct $x_2 - x_1$, a \textbf{minimal encoding} of $x_2$ in terms of $x_1$, such that the number of bits needed to specify the encoding scales linearly with $f(x_1, x_2)$.
We would need to construct a method for construction such \textit{minimal encodings} for each pairing of $(X, f)$.
See Sections~\ref{subsec:methods:compression} and~\ref{sec:datasets-and-distance-functions} for some examples.

@@

For example, given a dataset of genomic strings with the Levenshtein edit distance as the metric, let $x_1$ and $x_2$ be two strings in the dataset.
Let $x_2 - x_1$ be the shortest list of edits to convert $x_1$ into $x_2$.
The length of this edit list is equal to the Levenshtein edit distance i.e. $|x_2 - x_1| = f(x_1, x_2)$.
As in~\cite{berger2020levenshtein}, each such edit can be stored using $\mathcal{O}(\log m)$ bits where $m$ is the maximum length of any string in the dataset.
Thus, the total number of bits needed for each such encoding is $\mathcal{O} \big( f(x_1, x_2) \big)$.

Hence, for $|X| < \infty$,
\begin{align*}
    H (C) &\leq H(c) + \sum_{x \in C} H(x - c) \\
    &\leq H(c) + |C| \cdot H(y - c) \\ \\
    \text{where} \ y &= \argmax_{x \in C} \ f(x, c) \text{, or equivalently} \ f(y, c) = r.
\end{align*}

@@

We have two choices for encoding the instances in a cluster:
either each instance can be encoded in terms of the cluster center, 
or each child-cluster can be encoded in terms of the child-center and the child-centers encoded in terms of the cluster-center.

Thus the Shannon Entropy of a cluster encoding is given by
\begin{align}
    \label{eq:hierarchical-shannon-entropy}
    H(C) &\leq H(c - c_{parent}) + min \begin{cases}
        |C| \cdot H(y - c) \\
        H(C_{left}) + H(C_{right})
    \end{cases}
\end{align}

@@

\subsection{Scaling Behavior of Cluster Radii}
\label{subsec:methods:radii-scaling-behavior}

We show that the radii of clusters are guaranteed to decrease after, at most, every $d$ recursive applications of Partition.
The Partition algorithm is described in Section~\ref{subsec:methods:partition}.

The Manifold Hypothesis indicates that the instances in the dataset follow a $d$-dimensional distribution with $d \ll D$ where $D$ is the dimensionality of the embedding space.
We can measure the dimensionality of this low-dimensional distribution using the local fractal dimension of the dataset.
Thus, we can describe this distribution by choosing some set of $d$ mutually orthogonal axes.

Let $2r$ be the maximum pairwise distance among the instances in the dataset.
In the worst case, e.g. with the distribution shaped like a $d$-sphere, our chosen axes will be such that $2r$ is the maximum pairwise distance \textit{along every axis}.
Such a distribution would produce a balanced clustering.

Partition will select a maximally distant pair of points to use as poles, i.e. it will choose one of the $d$ axes along which to split the cluster into two children.
After one Partition, the maximum pairwise distance along that axis will be bounded above by $r$.
The next recursive Partition will select another of the $d$ axes.
Thus, after at most $d$ Partitions, the maximum pairwise distance along each axis will be bounded above by $r$.
The overall maximum pairwise distance, \textit{i.e. not restricted to be along one axis}, will be bounded above by $r \sqrt{2}$, for example, between two instances that lie at the extrema of different axes.

@@

Note that, in practice, we never see a balanced clustering.
Instead, Partition produces unbalanced trees due the varying density of sampling in different regions of the manifold.
Further, the cluster radii decrease by a factor much larger than $\frac{1}{\sqrt{2}}$ and they do so every one or two partitions rather than after $d$ partitions.
See Section~\ref{sec:results} for results on several datasets.

@@

In the worst case, we would have a balanced clustering.
Starting with cluster $C$ and applying $d$ recursive partitions would give us $2^d$ descendants with $|C| \cdot 2^{-d}$ instances each.
If cluster $C$ has radius $r$ then each of the $2^d$ descendants will have radius $\frac{r}{\sqrt{2}}$.

In the general case, let $T(C, d)$ be the subtree rooted at $C$ with depth $d$ and $J$ leaves.
Since we cannot partition a cluster containing only one instance, we can bound $J$ and $d$, i.e. $J \leq 2^d \leq |C|$.
Let the leaves be indexed with integers $j \in [1, J]$.
Let the center of each child-cluster be encoded in terms of the center of its parent-cluster.
Therefore, $H(T)$, the Shannon Entropy of $T(C, d)$, is the sum of all encodings of child-centers in terms of their parent-centers.

Thus, the second case in Equation~\ref{eq:hierarchical-shannon-entropy} becomes:
\begin{align*}
    & H(T) + \sum H(C_j) \\
    = \ & H(T) + \sum |C_j| \cdot H(y_j - c_j)
\end{align*}

From the analysis of the scaling behavior of cluster-radii in Section~\ref{subsec:methods:radii-scaling-behavior}, we have $H(y_j - c_j) \leq \frac{1}{\sqrt{2}} H(y - c)$.
Thus, we continue:
\begin{align*}
    \leq \ & H(T) + \frac{1}{\sqrt{2}} \cdot \sum |C_j| \cdot H(y - c) \\
    = \ & H(T) + \frac{1}{\sqrt{2}} \cdot |C| \cdot H(y - c)
\end{align*}

So we rewrite Equation~\ref{eq:hierarchical-shannon-entropy} as
\begin{align}
    \label{eq:hierarchical-shannon-entropy-2}
    H(C) &\leq H(c - c_{parent}) + \min \begin{cases}
        |C| \cdot H(y - c) \\
        |C| \cdot H(y - c) \cdot \frac{1}{\sqrt{2}} + H(T)
    \end{cases}
\end{align}

@@

For low depths in the tree, we expect that using subtrees to encode the instances will lead to great savings in the number of bits used.
As we continue deeper into the tree, we would start seeing diminishing returns for using subtrees.
Eventually, either the two cases will equalize or the first case will become the better choice.
At such a depth in the tree, we can encode the instances in a cluster using the cluster center rather than the subtree.
The center of that cluster can be encoded in terms of the center of its parent, and so on back up the tree to the root.

Notably, such a depth need not be uniform across the tree.
Indeed, the depth can adapt to the variations in sampling density and local fractal dimension in different regions of the manifold.

### 2021-04-07 sections/3-methods.tex
typos, reorganization, and more

@@

Exact partition costing $\mathcal{O}(n^2)$ vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.

% latex does not seem to like math symbols in this heading
\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

@@

\subsection{k-Nearest Neighbors Search}
\label{subsec:methods:knn-search}

@@

@@

Encode and Decode algorithms.
These will be specific for each dataset and/or distance function.

>>>>

Write about methods and provide asymptotic complexity analysis of each method.

@@

Exact partition with $\mathcal{O}(n^2)$ cost vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.
Building the exact tree costs $\mathcal{O}(n^2 \log n)$ vs approximate tree for $\mathcal{O}(n \log n)$.

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

@@

Complexity is the same as in CHESS.

\subsection{\texorpdfstring{$k$}{k}-Nearest Neighbors Search}
\label{subsec:methods:knn-search}

@@

Complexity is some factor times CHESS-search.

@@

Encode and Decode algorithms.
These could be, and probably will be, different for each $(X, f)$ pairing.

Asymptotic complexity analysis for:
\begin{enumerate}[i.]
    \item encode (novel contribution)
    \item serialize
    \item compress
    \item decompress
    \item deserialize
    \item decode (novel contribution)
\end{enumerate}

### 2021-04-07 sections/6-discussion.tex
hierarchical extension of 2020 Berger et. al. paper
typos, reorganization, and more

@@

More distance functions. For example, maximal-common-subgraph for molecular structures.

TODO: Provide link to source code on github.

>>>>

Is it the case that the optimal clusters for compression are also the optimal clusters for accelerated search?

What kind of trade-offs can we offer for space-savings of compression vs time-savings of accelerated-search in the compressed space?

@@

More distance functions. For example, Tanimoto Distance using maximal-common-subgraph for molecular structures.

Better methods for constructing minimal encodings, perhaps using domain expertise.
